---
title: "EnvImpactEval"
output: github_document
#output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Usage 

### The Baviaanskloof dataset
We will use the Baviaanskloof dataset created by  [del Río-Mena et al., 2021](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0243020), which is included in the \code{EnvImpactEval} package, to demonstrate the restoration impact evaluation. The dataset contains the sites of large-scale spekboom revegetation, and indicates in which year the revegetation was undertaken. We here select the sites revegetated in 2012 that were not subject to lifestock grazing exclusion to assess the effectiveness of the restoration actions. 

```{r, warning=FALSE, message=FALSE}
library(EnvImpactEval)

data(baviaanskloof)
baviaanskloof <- unwrap(baviaanskloof)

plot(baviaanskloof, "Planting_date")

year <- 2012
selected_sites <- baviaanskloof[baviaanskloof$Planting_date==year & baviaanskloof$Lifestock_exclusion==0]
other_sites <- baviaanskloof[baviaanskloof$Planting_date!=year | baviaanskloof$Lifestock_exclusion==1]
```


### Defining the spatial reference frame and the candidate control pixels
The \code{EnvImpactEval} package requires the user to define the spatial parameters at which the analysis will be performed. For now, only a raster-based framework is implemented. The user must provide the Coordinate Reference System of the raster (by default the UTM zone of the site's centroid), the spatial resolution (which is typically linked to the spatial resolution of EO datasets used in the analysis), and the parameters that define from which area control pixels for a control-impact or before-after-control-impact will be selected.

We here choose to restrict the search area for potential control units to a bounding box with a buffer of 3 km around the Baviaanskloof project area. Furthermore, in order to avoid undesired effects of spatial misregistration or adjacency effects, we exclude as control units those pixels in a radius of 150 m around the selected impact sites. Since we will use metrics derived from Landsat data to quantify restoration effectiveness, we selected a corresponding spatial resolution of 30 as the basis of our analysis. We can now plot a map showing the impact pixels (in green), the potential control pixels (in grey), and the excluded pixels (in white). Notice that the map shows UTM coordinates, as this was by default selected as the Coordinate Reference System. 

```{r}
refRas <- ref_rast(baviaanskloof, resolution=30, buffer=3000, round_coords=-2)
matchingCands <- matchCandidates(selected_sites, refRas, excludeBufferOut=150, excludeOther = other_sites)
plot(matchingCands)
plot(project(selected_sites, refRas), add=TRUE)
```

### Selecting the evaluation metrics




NDVI/SAVI







