---
title: "BACI evaluation Baviaanskloof"
author: Jasper Van doninck, Trinidad del Río-Mena, Wieteke Willemen, Wietske Bijker
output: 
  github_document:
    toc: true
    toc_depth: 1
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Baviaanskloof dataset
We here use a part of the Baviaanskloof dataset created by [del Río-Mena et al., 2021](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0243020), included in the `EnvImpactEval` package, to demonstrate a before-after-control-impact (BACI) evaluation using the package. The Baviaanskloof dataset contains the sites of large-scale spekboom (*Portulacaria afra*) revegetation in the semi-arid Baviaanskloof Hartland Bawarea Conservancy, South-Africa. Attributes to the spatial data indicate in which year the revegetation was undertaken by the different restoration organizations active in the area (Commonland, Living Land, Grounded) and whether lifestock grazing exclusion was implemented as a restoration activity or not. As an example, we here select the sites revegetated in 2012 that were not subject to lifestock grazing exclusion.

```{r, warning=FALSE, message=FALSE}
library(EnvImpactEval)
data(baviaanskloof)
baviaanskloof <- unwrap(baviaanskloof)

year <- 2012
selected_sites <- baviaanskloof[baviaanskloof$Planting_date==year & baviaanskloof$Lifestock_exclusion==0]
other_sites <- baviaanskloof[baviaanskloof$Planting_date!=year | baviaanskloof$Lifestock_exclusion==1]
plot(baviaanskloof, "Planting_date")
lines(selected_sites, lwd=2)
```

# Spatial reference and the candidate control pixels
The `EnvImpactEval` package requires the user to define the spatial parameters at which the analysis will be performed. This reference will define the spatial units of the analysis and the neighbourhood around the impact units in which we want to search for control units. For now, only a raster (pixel)-based framework is implemented, a vector-based framework is planned for a future release.  

Using the `ref_rast` function, the user can easily define this spatial reference raster based on the input data of the intervention to be evaluated (in our example the Baviaanskloof revegetation polygons for the year 2012). The spatial resolution of the reference raster will typically be linked to the spatial resolution of the (earth observation) datasets used in the analysis and we here set this to 120 meters. 
In order to limit data download and processing time we restrict the search area for potential control units to a bounding box with a buffer of 5 km around the sites reforested in 2012, but the user can choose to set this neighbourhood in function of their knowledge of their study sites.
The user can also specify the Coordinate Reference System (crs) of the reference raster. In our example we don't provide the crs parameter, in which case the UTM zone of the Baviaanskloof reference dataset will be used. 

```{r}
refRas <- ref_rast(selected_sites, resolution=120, buffer=5000, round_coords=-2)
```

The `matchCandidates` function is then used to further specify the impact units (pixels) and candidate control units that will be used in the BACI analysis. 
In order to avoid undesired effects of spatial misregistration or adjacency effects (the area just outside a restoration or conservation intervention site may be subject to positive or negative effects from the intervention), the user can exclude pixels in a buffer within or around the impact sites. In the case of an outside buffer this is in our example done with the argument `excludeBufferOut=240m`. We obviously also exclude the areas restored in the other years as control pixels with the `excludeOther` argument. 

```{r}
matchingCands <- matchCandidates(selected_sites, refRas, excludeBufferOut=150, excludeOther = other_sites)
```

We can now plot a map showing the impact pixels (in green), the potential control pixels (in grey), and the excluded pixels (in white). 
Notice that the map shows UTM coordinates (as this was by default for the `crs` argument in `ref_rast` function), rather than the geographic coordinates of the Baviaanskloof reference dataset. In general, the `EnvImpactEval` package will automatically check the crs of used spatial datasets and reproject when required.

```{r}
plot(matchingCands)
selected_sites_proj <- project(selected_sites, refRas)
lines(selected_sites_proj)
```

# Evaluation metric

We here chose Landsat Normalized Difference Vegetation Index (NDVI) as the metric to use in the impact evaluation. NDVI is indicative of vegetation "greenness" and could therefore inform us whether a large scale revegetation effort was successful. The Landsat programme has been running for several decades and is therefore a useful data source for impact evaluation of past interventions. Landsat data are available from several cloud platforms. We here use Microsoft's Planetary Computer, which allows querying with the SpatioTemporal Asset Catalog (STAC) language. 

```{r}
stac_endpoint <- as.endpoint("PlanetaryComputer")
stac_collection <- as.collection("Landsat", stac_endpoint)
```

The timeframes over which conservation or restoration impacts are assessed will depend on the application and study site. Assuming we want to evaluate the impact of the Baviaanskloof revegetation from the 10-years periods before and after the intervention year, we can first generate yearly NDVI time series (these steps can take a few minutes depending on the defined size and spatial resolution of the reference raster):
```{r}
vi_before <- VI_composite(endpoint=stac_endpoint, 
                         collection=stac_collection, 
                         VI="NDVI", 
                         spatRef = refRas, 
                         years = seq(year-10, year-1, 1), 
                         months = 3:5, 
                         maxCloud=60)

vi_after <- VI_composite(endpoint=stac_endpoint, 
                         collection=stac_collection, 
                         VI="NDVI", 
                         spatRef = refRas, 
                         years = seq(year+1, year+10, 1), 
                         months = 3:5, 
                         maxCloud=60)
```
In the above code, we specified that we want to make a pixel-based NDVI composite image for each year, using for each pixel and each year the median value from the images acquired in the months March until May that have a cloud cover below 60%.

We then calculate simple time series metrics from these two 10-year time series: the average value and trend (change in NDVI/year) over the time series. More advanced time series analysis methods will be added in a later version of the `EnvImpactEval` package. 

```{r}
avgtrend_before <- calc_ts_metrics(vi_before)
avgtrend_after <- calc_ts_metrics(vi_after)

plot(avgtrend_before, main=c("NDVI average, before", "NDVI trend, before"))
plot(avgtrend_after, main=c("NDVI average, after", "NDVI trend, after"))
```

# Control-Impact Matching

In impact assessment using control-impact (of before-after-control-impact) schemes, it is crucial that the control units are carefully selected so that their characteristics only differ from those of the impact group by the received "treatment" (in our case a restoration intervention). This is in the `EnvImpactEval` package done by specifying a set of covariates to be used in the matching analysis.  Ideally, users should include “all covariates likely to impact both the selection to the treatment and the outcome of interest” ([Schleicher et al.](https://onlinelibrary.wiley.com/doi/abs/10.1111/cobi.13448)). 

For the Baviaanskloof example, we use a limited number of matching covariates. Because we restricted the search window for control units to a relatively small area around the impact sites, we assume that climatic variation will be small and mostly induced by local topography. Topographic slope and aspect may also determine whether or not a site is selected for revegetation or determine the success of the revegetation intervention. This terrain data can be extracted from the NASA Digital Elevation Model (DEM) based on SRTM data using the `nasadem` function. Unless specified otherwise, the DEM will be obtained from Microsoft's Planetary Computer. The user can also specify the terrain derivatives to be calculated (by default the slope and the aspect), and whether the aspect should be transformed to "northness" and "eastness" by applying the cosine and sine function on it (default) or not.  

```{r}
dem <- nasadem(refRas)
plot(dem)
```

We also included land cover as a matching covariate. This will ensure that control pixel are selected from the same landcover type (i.e. "rangeland"). Note that in the current implementation the 9-class land use land cover layer from Impact Observatory is used. This product is first generated for the year 2017, so we assume no change in land cover change between the intervention year (2012) and 2017. Given that terrain revegetated with spekboom will still be classified as "rangeland" this is not an issue in this example, but it may be a problem for other situations.

```{r}
landcover <- lulc(refRas, year=year)
plot(landcover)
lines(selected_sites_proj)
```

Distance to roads or settlements can explain the anthropogenic pressure on an area. The `EnvImpactEval` package therefore includes functions to calculate the distance to the nearest road (`osm_distance_roads`) or nearest settlement (`osm_distance_places`) based on OpenStreetMap data. For this example, we include the distance to the nearest road of OSM category "track" or higher as an environmental covariate in the matching analysis.

```{r}
roadsDist <- osm_distance_roads(refRas, values="track+")
plot(roadsDist)
lines(selected_sites_proj)
```

In addition to parameters described above, we also included as matching covariates the average and trend of the selected vegetation index in the 10-year period before the restoration intervention. This way, control pixels will be selected that have undergone a similar temporal trajectory as the impact pixels in the years before intervention. We then perform the control-impact matching using propensity score matching (default method), and select 10 control pixels for each impact pixel with replacement (a control pixel can be paired with several impact pixels). This step can be performed interactively by setting the parameter `eval=TRUE`, which will display matching results and plots after which the matching can be accepted or rejected.

```{r}
matchingLayers <- list(dem, landcover, roadsDist, avgtrend_before)
matches <- matchCI(matchingCands, rast(matchingLayers), eval=TRUE, ratio=10, replace=TRUE)
```

From the output of the matching analysis, we can see that the Absolute Standardized Mean Difference for all covariates is below (or marginally above) the threshold of 0.1. We therefore accept this matching and use it as an input to the BACI evaluation.

# Before-After-Control-Impact evaluation

To tun the BACI analysis, we first need to extract the values of the evaluation metrics (in this case the average over the 10-year time series before and after intervention) for the control-impact pairs obtained in the matching. When running the BACI analysis, we specify a spatial reference to which the pixel-based results will be written. 

```{r}
baci_input <- extractMatches(matches, before=subset(avgtrend_before, "average"), after=subset(avgtrend_after, "average"))
baci_results <- BACI_contrast(baci_input, SpatRef=crop(refRas, selected_sites_proj))
```

The results of the BACI analysis can be viewed by printing the data.table associated to the output object. The data.table shows, for each impact unit, the BACI contrast and p-values. BACI contrast values smaller than zero indicate a positive effect of the revegetation intervention. 

```{r}
baci_results$data
```

The BACI results can also be plotted on a map, here with non-significant BACI contrast masked out. If desired, the pixel-based results can now be aggregated to different spatial levels, e.g., the revegetation site.
```{r}
plot(baci_results$spat["contrast"], 
     range=baci_results$spat["contrast"] |> minmax() |> abs() |> max() |> rep(2)*c(-1,1), 
     col=hcl.colors(50, palette = "RdYlBu", rev = TRUE))
plot(as.numeric(baci_results$spat["p_value"]<0.05), 
     col=data.frame(value=c(0,1), color=c("grey95", NA)), 
     add=TRUE, 
     legend=FALSE)
lines(selected_sites_proj)
```


