---
title: "wildlife_circles_Finland"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spatialBACI)
library(terra)
```

```{r, include=FALSE}
terra::terraOptions(memfrac = 0.9)
```


# Wildlife circles dataset

To illustrate how the `spatialBACI` package can be used with pre-defined monitoring sites, we use the `wildlife_circles` dataset, which is a simplified subset of the data used by [Terraube et al., 2020](https://www.nature.com/articles/s41467-020-16792-7). The dataset is a simplification of the Finnish wildlife triangles monitoring scheme ([Helle et al., 2016](https://cdnsciencepub.com/doi/10.1139/cjfr-2015-0454)), where each circle is designed to pass through the three vertices of the triangle made up of transects of 4 km. The `wildlife_circles` dataset provides the outline of the circles as polygons with a generic "ID" assigned, and indicates for each circles whether it is located in a protected area (according to the [World Database on Protected Areas](https://www.protectedplanet.net/en/thematic-areas/wdpa?tab=WDPA)) or not (value of 1/0 in attribute "PA").


```{r, warning=FALSE, message=FALSE}
data(wildlife_circles)
wildlife_circles <- unwrap(wildlife_circles)

wildlife_circles
plot(centroids(wildlife_circles), "PA", )
```

In [Terraube et al., 2020](https://www.nature.com/articles/s41467-020-16792-7), the authors want to match wildlife circles inside protected areas to control sited outside protected areas using six matching covariates:

- Latitude
- Longitude
- Distance to the closest settlement
- Terrain ruggedness
- Human population density
- Percentage forest cover

We will here show how to perform such analysis using the `spatialBACI` package and other common R packages.

# Matching covariates

Extracting **longitude** and **latitude** of the centroids of polygons in a SpatVector object, as is the case here, is rather straightforward using the `terra::as.data.frame()` functions of the `terra` package, on which the `spatialBACI` package builds. The following returns a data frame with for each wildlife circle the original "ID" and "PA" columns, and additionally columns "x" and "y" indicating the longitude and latitude of the centroid of the wildlife circle.

```{r}
LonLat <- centroids(wildlife_circles) |> as.data.frame(geom="XY")
```

**Distance to the closest settlement** can be obtained using `osm_distance_places` function, where we define settlements OpenStreetMap features of the category "village" or above. We set the timeout argument to 100 (seconds) to allow retrieval of all settlements in Finland.

```{r, warning=FALSE}
dist_village <- osm_distance_places(x=wildlife_circles, values="village+", timeout=100, osm_bbox = "Finland")
```

**Terrain ruggedness** was by Terraube et al. defined as the standard deviation of the elevation within the wildlife triangles, and was included because large carnivores were expected to reach higher densities in more rugged terrain. Because Finland is mostly to the north of the area covered by the SRTM DEM, we will derive the terrain ruggedness from the ALOS DEM freely available in the Planetary Computer STAC catalog. We specify that we want to use the original 30m resolution in x and y of the ALOS DEM.

```{r}
dem_finland <- dem(x=wildlife_circles,
                 v="elevation",
                 dem_source=list(endpoint="https://planetarycomputer.microsoft.com/api/stac/v1",
                                 collection="alos-dem",
                                 assets="data"),
                 dx=30, dy=30)
```

There is no specific function in the `spatialBACI` package to extract **human population density**. However, data in raster format stored locally or accessible online can also easily be integrated. In this example, we access the [WorldPop](https://hub.worldpop.org/) population density of Finland for the year 2016 (the year used in Terraube et al., 2020) at 1km resolution, found at <https://hub.worldpop.org/geodata/summary?id=41175>. 
```{r}
popdens_url <- "https://data.worldpop.org/GIS/Population_Density/Global_2000_2020_1km/2016/FIN/fin_pd_2016_1km.tif"
```

**Percentage forest cover** was calculated by Terraube et al. (2020) from the CORINE Land Cover (CLC) dataset for the year 2012. We downloaded the CLC dataset for the year 2012 from the Finnish Environment Institute (SYKE) website <https://www.syke.fi/en/environmental-data/downloadable-spatial-datasets#corine-land-cover> (CORINE Land Cover 2012, 20 m geotiff (zip) (latest update 30.9.2014)) and provided the extracted data with the `spatialBACI` package. Extracting the forest fraction over each wildlife circle requires knowledge of the values under which the forest classes are stored (in this case values 22 through 29), and a custom function `forestFrac_function`. 

```{r}
clc_filename <- system.file("extdata/clc2012_fi20m.tif", package="spatialBACI")
forestFrac_function <- function(x){sum(x %in% 22:29)/length(x)}
```

The `collate_matching_layers()` function is then designed to convert all these matching covariates in different formats (data.frame, SpatVector, online or locally stored rasters). For this, we have to provide as inputs to `collate_matching_layers()` the SpatVector file with the candidate control and impact sites (`wildlife_circles`), and a list of the matching variables. For matching variables in raster format, additional information must be provided on how the raster data will be summarized over the vector units of analysis. In this case, the raster dataset must be provided as a named list element "data", and additional named list elements specify the summarizing function. For population density, we want to calculate the interpolated value of the four 1km pixels adjacent to the centroids of the wildlife circles, which can be used by adding the list element `method="bilinear"`. Surface roughness can be derived from the DEM by setting the list element `fun=sd` (and `na.rm=TRUE` to ignore occasional missing values in the DEM). Finally, for formula to derive forest fraction is linked to the CLC raster dataset. This step may take a few minutes, as the DEM data will in this step be downloaded and processed.

```{r, warning=FALSE}
vars_list <- list(LonLat, 
                  dist_village, 
                  list(data=clc_filename,
                       fun=forestFrac_function),
                  list(data=popdens_url,
                       method="bilinear"),
                  list(data=dem_finland,
                       fun=sd, 
                       na.rm=TRUE)
                  )

matching_input_vector <- collate_matching_layers(wildlife_circles, vars_list=vars_list)
```

In the resulting object, the column names of the data.table referring to the raster objects are simply taken from the layer names of the corresponding file. For clarity, we can update the column names (though they should not contain spaces). 

```{r}
matching_input_vector
names(matching_input_vector$data)[3:8] <- c("easting", "northing", "dist_villages", "forest_fraction" ,"pop_density", "terrain_roughness")
```

# Control-impact matching

To match one wildlife circle outside protected area to each site within a protected area without replacement using the collated matching covariates, we simply provide this resulting object to the `matchCI()` function. We must provide the column names defining the unique ID of each spatial unit of analysis (`colname.id="ID"`) and the treatment (`colname.treatment="PA"`). In this example, we use 1:1 matching without replacement. 

```{r}
matching_output_vector <- matchCI(matching_input_vector,
                                  colname.id="ID", colname.treatment="PA",
                                  ratio=1, replace=FALSE, method="nearest")
```

Matching results can now be evaluated using `evaluate_matching()`, after which the user can decide to continue their analysis with the matched dataset or repeat matching with different parameters. 

```{r}
evaluate_matching(matching_output_vector)
```

